---
title: "Elena's nascent RNAseq analysis"
author: "Jose Alejandro Romero Herrera"
date: '`r Sys.Date()`'
knit: (function(inputFile, encoding) { 
      rmarkdown::render(inputFile,
                        encoding=encoding,
                        output_format='all',
                        output_dir='../reports/')})
output:
  # To create PDF report, uncomment below
  #pdf_document:
  #  toc: yes
  html_document:
    number_sections: yes
    theme: yeti
    toc: yes
    toc_float: yes
    df_print: paged
    dev: png
---

```{r knitr, include = FALSE}
DOCNAME = knitr::current_input()
knitr::opts_chunk$set(autodep        = TRUE,
                      cache          = FALSE,
                      cache.path     = paste0("cache/", DOCNAME, "/"),
                      cache.comments = TRUE,
                      echo           = TRUE,
                      error          = FALSE,
                      fig.align      = "center",
                      fig.path       = paste0("../reports/figures/", DOCNAME, "/"),
                      fig.width      = 10,
                      fig.height     = 8,
                      message        = FALSE,
                      warning        = FALSE)
```

# Introduction

This is a clean version of Elena's RNAseq analysis of the effect of Oct4 knockdown after pERK activation.

# Load packages
```{r, warning=FALSE, message=FALSE, error= FALSE}
library(tidyverse)
library(DESeq2)
library(RColorBrewer) #colour scheme for hierarchical clustering heatmap
library(ggdendro)
library(ggrepel)
library(reshape2)
library(apeglm) #for LFC shrinkage in DESEq2
library(genefilter) #top var genes
library(pheatmap)
library(readxl)
library(xlsx)
library(biomaRt)
```

Create results folder for this notebook
```{r}
results_folder <- "../results/complete_deseq2_analysis/"

if(!file.exists(results_folder)){
  dir.create(results_folder)
}
```

# Load data
```{r}
data_path <- "../data/counts_gene-id_gene.txt"
data <- read.delim(data_path, as.is=T)
head(data)
```

Separate counts from gene metadata
```{r}
gene_metadata <- data[,1:6]
rownames(data) <- data$Geneid
data <- data[,-c(1:6)]
```

## Prepare gene names and ids
Getting gene names from the mouse ensembl data base using Biomart
```{r}
genes_biomart <- data.frame(ensembl_gene_id = as.character(gene_metadata$Geneid))
```


```{r}
getGeneAnnotationsMM = function(d="./",force=F){
  
  mart <- useMart(
    host="https://www.ensembl.org", biomart = 'ENSEMBL_MART_ENSEMBL',
    dataset = 'mmusculus_gene_ensembl')
  
  annot <- biomaRt::getBM(
    attributes = c(
      "ensembl_gene_id",
      "external_gene_name",
      "description"),
    mart = mart)
  
  return(annot)
}

gene_annot <- getGeneAnnotationsMM()
genes_biomart <- merge(genes_biomart, gene_annot, by.x = "ensembl_gene_id", sort = F, all.x = T)
head(genes_biomart)
```

Replacing missing gene names for original ensembl id
```{r}
genes_biomart$external_gene_name[is.na(genes_biomart$external_gene_name)] <- genes_biomart$ensembl_gene_id[is.na(genes_biomart$external_gene_name)]
```

Checking duplicates
```{r}
# Check for gene id and name duplicates
check_duplicates <- function(x, unique = T, show = F) {
  duplicates <- sum(duplicated(x))
  
  if (duplicates == 0) {
    print("There are no duplicates")
    invisible(x)
  }
  else{
    print(paste0("There are ",duplicates," duplicated IDs"))
    if (show) {
      print(table(x[duplicated(x)]))
    }
    if (unique) {
      print(paste0("Making them unique"))
      invisible(make.names(x, unique = T))
    }
    else{
      invisible(x)
    }
  }
}

check_duplicates(genes_biomart$ensembl_gene_id, unique = F)
check_duplicates(genes_biomart$external_gene_name, unique = F, show = T)
```

## Prepare sample metadata
Extract information (metadata) from sample column names
```{r}
names = colnames(data)
names = sapply(names,function(x) gsub("X8h","X_8h",x))
p1 = "(.*)_(.*)_(.*)_(.*)_(.*)_uniq"
colData <- data.frame(str_match(names,p1),stringsAsFactors = F)
colnames(colData) = c("Sample","Dox","Time","Tam","Rep","Rep2")
colData$SampleName = colnames(data)
colData$TimePt = as.numeric(str_match(colData$Time,"^(.*)h")[,2])

# samples to fix: the Tam field should be empty
# and the values there should move to Dox
## Jose: This does not seem to apply to the 8h time point, i believe it is for the ones before the 2nd treatment
idx = colData$Dox=="X"
colData[idx,"Dox"] = colData[idx,"Tam"]
colData[idx,"Tam"] = NA

colData$SampleNameShort = with(colData,paste(Dox,Tam,Time,Rep2, sep="_"))
colData$ShortLabel = with(colData,paste(Dox,Tam,Time, sep="_"))
colData$ShortLabelTime = with(colData,paste(Dox,Tam, sep="_"))
colData$ShortLabel = factor(colData$ShortLabel)
colData$ShortLabel = relevel(colData$ShortLabel, "H2O_NA_8h")
```


```{r}
head(colData)
```

# DESEQ pre-analysis
Removing rows with no counts
```{r}
filter <- rowSums(data) > 1
print(paste0("Removing ", nrow(data) - sum(filter), " rows"))
data <- data[filter,]
```

## Create DESEQ object
```{r}
#full condition
dds <- DESeqDataSetFromMatrix(data, colData, design = ~ ShortLabel)
dds <- DESeq(dds)
```

## Vst normalization
```{r}
vsd <- vst(dds)
```

### Principal component plot of the samples
PCA plot using the first two components
```{r}
pcaData <- plotPCA(vsd, intgroup=c("ShortLabel"), returnData=TRUE)
percentVar <- round(100 * attr(pcaData, "percentVar"))
```


```{r, PCA_plot}
ggplot(pcaData, aes(PC1, PC2, color=ShortLabel, label=ShortLabel)) +
  geom_point(size=3) + geom_text_repel() +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  coord_fixed() + theme_bw()
```


### Heatmap of the sample-to-sample distances
This plot shows how far away are each sample from each other. The darker the blue, the closer they are.
```{r distance_heatmap, fig.height=7, fig.width=7}
sampleDists <- dist(t(assay(vsd)))
sampleDistMatrix <- as.matrix(sampleDists)
rownames(sampleDistMatrix) <- vsd$ShortLabel
colnames(sampleDistMatrix) <- NULL
colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255)
pheatmap(sampleDistMatrix,
         clustering_distance_rows=sampleDists,
         clustering_distance_cols=sampleDists,
         col=colors)
```

# DE analysis
We have to do several DE analysis and create a list with all DE across conditions and timepoints
```{r}
LFC <- 1
alpha <- 0.05

filter_results <- function(result,LFC, alpha){
    result <- as.data.frame(result)
    result <- result[complete.cases(result),]
    result <- result[abs(result$log2FoldChange) > LFC & result$padj < alpha,]
    return(result)
}

result_list <- list()
```

## 0h comparisons
DOX_NA_8h vs H2O_NA_8h
The proper reference level is already establish
```{r}
resultsNames(dds)
res <- results(dds, name = "ShortLabel_DOX_NA_8h_vs_H2O_NA_8h")
result_list[["DOX_NA_8h_vs_H2O_NA_8h"]] <- filter_results(res, LFC, alpha)
```

## 2h comparisons
DOX_ETOH_2h vs H2O_ETOH_2h
H2O_4OHT_2h vs H2O_ETOH_2h
DOX_4OHT_2h vs H2O_ETOH_2h
```{r}
dds$ShortLabel <- relevel(dds$ShortLabel, ref = "H2O_ETOH_2h")
dds <- DESeq(dds)
resultsNames(dds)

res <- results(dds, name = "ShortLabel_DOX_ETOH_2h_vs_H2O_ETOH_2h")
result_list[["DOX_ETOH_2h_vs_H2O_ETOH_2h"]] <- filter_results(res, LFC, alpha)

res <- results(dds, name = "ShortLabel_H2O_4OHT_2h_vs_H2O_ETOH_2h")
result_list[["H2O_4OHT_2h_vs_H2O_ETOH_2h"]] <- filter_results(res, LFC, alpha)

res <- results(dds, name = "ShortLabel_DOX_4OHT_2h_vs_H2O_ETOH_2h")
result_list[["DOX_4OHT_2h_vs_H2O_ETOH_2h"]] <- filter_results(res, LFC, alpha)
```
## 8h comparisons
DOX_ETOH_8h vs H2O_ETOH_8h
H2O_4OHT_8h vs H2O_ETOH_8h
DOX_4OHT_8h vs H2O_ETOH_8h
```{r}
dds$ShortLabel <- relevel(dds$ShortLabel, ref = "H2O_ETOH_8h")
dds <- DESeq(dds)
resultsNames(dds)

res <- results(dds, name = "ShortLabel_DOX_ETOH_8h_vs_H2O_ETOH_8h")
result_list[["DOX_ETOH_8h_vs_H2O_ETOH_8h"]] <- filter_results(res, LFC, alpha)

res <- results(dds, name = "ShortLabel_H2O_4OHT_8h_vs_H2O_ETOH_8h")
result_list[["H2O_4OHT_8h_vs_H2O_ETOH_8h"]] <- filter_results(res, LFC, alpha)

res <- results(dds, name = "ShortLabel_DOX_4OHT_8h_vs_H2O_ETOH_8h")
result_list[["DOX_4OHT_8h_vs_H2O_ETOH_8h"]] <- filter_results(res, LFC, alpha)
```
## All DE genes for all conditions
```{r}
DE_ids <- unique(unlist(lapply(result_list, rownames), use.names = F))
```

# Save all results
In an excel file
```{r}
add_data <- function(name,data,workbook){
  df <- data[name]
  new_sheet <- createSheet(workbook, sheetName = name)
  addDataFrame(df, sheet = new_sheet)
}
```

```{r}
wb <- createWorkbook()
sheet_names <- names(result_list)

lapply(sheet_names, FUN = add_data, result_list, wb)
saveWorkbook(wb, file = paste0(results_folder, "DEG_all_comparisons.xlsx"))
```

# Heatmaps
```{r}
vsd <- assay(vsd)

vsd <- aggregate(t(vsd), by = list(colData$ShortLabel), mean)
vsd_names <- vsd[,1]
vsd <- vsd[,-1]
vsd <- data.frame(t(vsd))
colnames(vsd) <- vsd_names
```

## pERK activation only
```{r perk_activation_heatmap}
perk_activation <- vsd[DE_ids,c("H2O_NA_8h", "H2O_4OHT_2h", "H2O_4OHT_8h")]
                          
pheatmap(perk_activation, show_rownames = F, cluster_cols = F, scale = "row")
```

## pERK activation and OCT4 KO
```{r perk_activation_Oct4_heatmap}
oct4_downregulation <- vsd[DE_ids,c("H2O_NA_8h", "DOX_4OHT_2h", "DOX_4OHT_8h")]
                          
pheatmap(oct4_downregulation, show_rownames = F, cluster_cols = F, treeheight_row = 0, scale = "row")
```

## All
```{r all_heatmap}
heatmap_data <- vsd[DE_ids, c("H2O_NA_8h", "H2O_4OHT_2h", "H2O_4OHT_8h", "DOX_4OHT_2h", "DOX_4OHT_8h")]

annot_col <- data.frame(pErk = c("0h", "2h","8h","2h","8h"),
                        Oct4 = c("+", "+", "+","-","-"),
                        row.names = colnames(heatmap_data))

colnames(annot_col)[1] <- "pErk activation"

pheatmap(heatmap_data, show_rownames = F, cluster_cols = F, treeheight_row = 0, scale = "row", show_colnames = F, annotation_col = annot_col)
```

# Josh's requests
These plots were asked by Josh for a grant report

## Only UP regulated Genes in the pERK time course

```{r}
pERK_results <- result_list[c(3,6,4,7)]

test <- function(x){
 return(rownames(x[x$log2FoldChange > 0,]))
}

DE_ids_up <- unique(unlist(lapply(pERK_results,test), use.names = F))
```

```{r}
heatmap_data <- vsd[DE_ids_up, c("H2O_NA_8h", "H2O_4OHT_2h", "H2O_4OHT_8h", "DOX_4OHT_2h", "DOX_4OHT_8h")]

annot_col <- data.frame(pErk = c("0h", "2h","8h","2h","8h"),
                        Oct4 = c("+", "+", "+","-","-"),
                        row.names = colnames(heatmap_data))

colnames(annot_col)[1] <- "pErk activation"
```

```{r}
pheatmap(heatmap_data, show_rownames = F, cluster_cols = F, treeheight_row = 0, scale = "row", show_colnames = F, 
         annotation_col = annot_col, cluster_rows = T,
         main = "Up-regulated genes after pERK activation\nwith and without Oct4")
```

```{r pERK_up_heatmap_nocluster}
pheatmap(heatmap_data, show_rownames = F, cluster_cols = F, treeheight_row = 0, scale = "row", show_colnames = F, 
         annotation_col = annot_col, cluster_rows = F,
         main = "Up-regulated genes after pERK activation\nwith and without Oct4")
```

Divide clusters and select 2h and 8h cluster
```{r}
cluster_2h <- t(scale(t(heatmap_data[,c("H2O_NA_8h","H2O_4OHT_2h", "DOX_4OHT_2h")])))
cluster_2h <- hclust(dist(cluster_2h))

cluster_8h <- t(scale(t(heatmap_data[,c("H2O_NA_8h","H2O_4OHT_8h", "DOX_4OHT_8h")])))
cluster_8h <- hclust(dist(cluster_8h))
```

```{r pERK_up_2h_heatmap}
pheatmap(heatmap_data, show_rownames = F, cluster_cols = F, cluster_rows = cluster_2h,
         treeheight_row = 0, scale = "row", show_colnames = F, annotation_col = annot_col,
         main = "Up-regulated genes after pERK activation\nwith and without Oct4 (genes clustered by 2h)")
```

```{r pERK_up_8h_heatmap}
pheatmap(heatmap_data, show_rownames = F, cluster_cols = F,cluster_rows = cluster_8h,
         treeheight_row = 0, scale = "row", show_colnames = F, annotation_col = annot_col,
         main = "Up-regulated genes after pERK activation\nwith and without Oct4 (genes clustered by 8h)")
```

## Only DOWN regulated Genes in the pERK time course

```{r}
pERK_results <- result_list[c(3,6,4,7)]

test <- function(x){
 return(rownames(x[x$log2FoldChange < 0,]))
}

DE_ids_down <- unique(unlist(lapply(pERK_results,test), use.names = F))
```

```{r}
heatmap_data <- vsd[DE_ids_down, c("H2O_NA_8h", "H2O_4OHT_2h", "H2O_4OHT_8h", "DOX_4OHT_2h", "DOX_4OHT_8h")]

annot_col <- data.frame(pErk = c("0h", "2h","8h","2h","8h"),
                        Oct4 = c("+", "+", "+","-","-"),
                        row.names = colnames(heatmap_data))

colnames(annot_col)[1] <- "pErk activation"
```

```{r pERK_down_heatmap}
pheatmap(heatmap_data, show_rownames = F, cluster_cols = F, treeheight_row = 0, scale = "row", 
         show_colnames = F, annotation_col = annot_col,
         main = "Down-regulated genes after pERK activation\nwith and without Oct4")
```

```{r pERK_down_heatmap_nocluster}
pheatmap(heatmap_data, show_rownames = F, cluster_cols = F, treeheight_row = 0, scale = "row", 
         show_colnames = F, annotation_col = annot_col, cluster_rows = F,
         main = "Down-regulated genes after pERK activation\nwith and without Oct4")
```

Divide clusters and do only 2 and 8h
```{r}
cluster_2h <- t(scale(t(heatmap_data[,c("H2O_NA_8h","H2O_4OHT_2h", "DOX_4OHT_2h")])))
cluster_2h <- hclust(dist(cluster_2h))

cluster_8h <- t(scale(t(heatmap_data[,c("H2O_NA_8h","H2O_4OHT_8h", "DOX_4OHT_8h")])))
cluster_8h <- hclust(dist(cluster_8h))
```


```{r pERK_down_2h_heatmap}
pheatmap(heatmap_data, show_rownames = F, cluster_cols = F, cluster_rows = cluster_2h,
         treeheight_row = 0, scale = "row", show_colnames = F, annotation_col = annot_col,
         main = "Down-regulated genes after pERK activation\nwith and without Oct4 (genes clustered by 2h)")
```


```{r pERK_down_8h_heatmap}
pheatmap(heatmap_data, show_rownames = F, cluster_cols = F,cluster_rows = cluster_8h,
         treeheight_row = 0, scale = "row", show_colnames = F, annotation_col = annot_col,
         main = "Down-regulated genes after pERK activation\nwith and without Oct4 (genes clustered by 8h)")
```

# Session info

```{r session-info, cache = FALSE}
devtools::session_info()
```
