---
title: "DEA heatmap for RNAseq"
author: "Jose Alejandro Romero Herrera"
date: '`r Sys.Date()`'
knit: (function(inputFile, encoding) { 
      rmarkdown::render(inputFile,
                        encoding=encoding,
                        output_format='all',
                        output_dir='../reports/')})
output:
  # To create PDF report, uncomment below
  #pdf_document:
  #  toc: yes
  html_document:
    number_sections: yes
    theme: yeti
    toc: yes
    toc_float: yes
    df_print: paged
    dev: pdf
---

```{r knitr, include = FALSE}
DOCNAME = knitr::current_input()
knitr::opts_chunk$set(autodep        = TRUE,
                      cache          = FALSE,
                      cache.path     = paste0("cache/", DOCNAME, "/"),
                      cache.comments = TRUE,
                      echo           = TRUE,
                      error          = FALSE,
                      fig.align      = "center",
                      fig.path       = paste0("../reports/figures/", DOCNAME, "/"),
                      fig.width      = 10,
                      fig.height     = 8,
                      message        = FALSE,
                      warning        = FALSE)
```

# Introduction
In this script we will generate the heatmap plot for Elena's nascent RNAseq. First, the whole plot. Then we will separate the genes by the row annotation:
 - pErk mutant UP + pErk UP
 - pErk mutant UP + pErk DOWN
 - pErk mutant DOWN + pErk UP
 - pErk mutant DOWN + pErk DOWN
 
# Load libraries

```{r}
library(DESeq2)
library(tidyverse)
library(RColorBrewer) #colour scheme for hierarchical clustering heatmap
library(ggrepel) #labels for ggplot
library(ggdendro)
library(reshape2)
library(apeglm) #for MAplot function in DESEq2
library(genefilter) #top var genes
library(pheatmap)
library(readxl)
library(stringr)
```
Create results folder for this notebook
```{r}
results_folder <- "../results/heatmap/"

if(!file.exists(results_folder)){
  dir.create(results_folder)
}
```

# Loading initial data
```{r}
data_path <- "../data/counts_gene-id_gene.txt"
data <- read.delim(data_path,as.is=T)
```

Annotations of the data
```{r}
annot <- data[,1:6]
```

Counts. "-11" is to pick the 8h samples (which are the 12 last samples of the file)

```{r}
counts <- as.matrix(data[,(ncol(data)-11):ncol(data)])
```

Getting gene names from the mouse ensembl data base
```{r}
genes <- annot$Geneid
require(org.Mm.eg.db)
db <- org.Mm.eg.db
ensembl <- suppressWarnings(mapIds(db, keys=genes, keytype="ENSEMBL", column="SYMBOL",multiVals="first"))
genes <- data.frame(SYMBOL = ensembl, ENSEMBL = genes, stringsAsFactors = F)
genes$SYMBOL[is.na(genes$SYMBOL)] = genes$ENSEMBL[is.na(genes$SYMBOL)]

annot$GeneSymbol = genes$SYMBOL
```


Getting sample annotations for DEseq analysis. We need to extract information (metadata) from sample column names
```{r}
names = colnames(counts)
names = sapply(names,function(x) gsub("X8h","X_8h",x))
p1 = "(.*)_(.*)_(.*)_(.*)_(.*)_uniq"
colData <- data.frame(str_match(names,p1),stringsAsFactors = F)
colnames(colData) = c("Sample","Dox","Time","Tam","Rep","Rep2")
colData$SampleName = colnames(counts)
colData$TimePt = as.numeric(str_match(colData$Time,"^(.*)h")[,2])
```

samples to fix: the Tam field should be empty and the values there should move to Dox. This does not seem to apply to the 8h time point, i believe it is for the ones before the 2nd treatment
```{r}
idx = colData$Dox=="X"
colData[idx,"Dox"] = colData[idx,"Tam"]
colData[idx,"Tam"] = NA

colData$SampleNameShort = with(colData,paste(Dox,Tam,Time,Rep2, sep="_"))
colData$ShortLabel = with(colData,paste(Dox,Tam,Time, sep="_"))
colData$ShortLabelTime = with(colData,paste(Dox,Tam, sep="_"))
```

# DOX_4OHT_vs_H2O_4OHT comparison

## Create colData
```{r}
colData$condition = factor(colData$ShortLabel, levels = c("H2O_4OHT_8h","H2O_ETOH_8h","DOX_ETOH_8h","DOX_4OHT_8h"))
colData$Dox = factor(colData$Dox, levels = c("H2O","DOX"))
colData$Tam = factor(colData$Tam, levels = c("ETOH", "4OHT"))
```

## DESeq analysis ###

```{r}
dds <- DESeqDataSetFromMatrix(counts, colData, design = ~ condition)
```

Adding the extrated annotation that we did at the beginning and add it to the information that we just retrieved

```{r}
mcols(dds) = cbind(annot,mcols(dds))
```

Pre-filtering the dataset
```{r}
nrow(dds)
dds <- dds[rowSums(counts(dds)) > 10, ]
nrow(dds)
```

## Results 
```{r}
dds <- DESeq(dds)
res <- results(dds, name = "condition_DOX_4OHT_8h_vs_H2O_4OHT_8h")
res <- as.data.frame(res)
res$gene_name <- mcols(dds)$Geneid
res[["padj"]][is.na(res[["padj"]])]<-0.99
res[["log2FoldChange"]][is.na(res[["log2FoldChange"]])]<-0
```

Extract Up and down regulated genes
```{r}
up_dox_tam_vs_h2o_tam<-filter(res,log2FoldChange>1, padj<0.05)$gene_name
down_dox_tam_vs_h2o_tam<-filter(res,log2FoldChange<(-1),padj<0.05)$gene_name

up_and_down_dox_tam_vs_h2o_tam <- filter(res, abs(log2FoldChange)> 1, padj < 0.05)
```

# H2O_ETOH as reference

Modify colData
```{r}
colData$condition = factor(colData$ShortLabel, levels = c("H2O_ETOH_8h","H2O_4OHT_8h","DOX_ETOH_8h","DOX_4OHT_8h"))
colData$Dox = factor(colData$Dox, levels = c("H2O","DOX"))
colData$Tam = factor(colData$Tam, levels = c("ETOH", "4OHT"))
```

## DESeq analysis
```{r}
dds <- DESeqDataSetFromMatrix(counts, colData, design = ~ condition)
```

Adding the extrated annotation that we did at the beginning and add it to the information that we just retrieved
```{r}
mcols(dds) = cbind(annot,mcols(dds))
```

Pre-filtering the dataset
```{r}
nrow(dds)
dds <- dds[rowSums(counts(dds)) > 10, ]
nrow(dds)
```

### DE analysis 4OHT vs Control

Get Up and down regulated pErk regulated genes
```{r}
dds <- DESeq(dds)
res <- results(dds, name = "condition_H2O_4OHT_8h_vs_H2O_ETOH_8h")
res <- as.data.frame(res)
res$gene_name <- mcols(dds)$Geneid
res[["padj"]][is.na(res[["padj"]])]<-0.99
res[["log2FoldChange"]][is.na(res[["log2FoldChange"]])]<-0

up<-filter(res,log2FoldChange>1, padj<0.05)
down<-filter(res,log2FoldChange<(-1),padj<0.05)

up_and_down_tam_vs_etoh <- filter(res, abs(log2FoldChange)> 1, padj < 0.05)
up_and_down_tam_vs_etoh$DE <- "Up"
up_and_down_tam_vs_etoh$DE[up_and_down_tam_vs_etoh$log2FoldChange < (-1)] <- "Down"
```

# Heatmap of genes regulated by pErk

```{r}
#mat <- assay(normTransform(dds))
mat <- assay(rlog(dds))
rownames(mat) <- mcols(dds)$Geneid
mat  <- mat[up_and_down_tam_vs_etoh$gene_name,]
#mat  <- mat - rowMeans(mat)
mat <- mat[,c("H2O_8h_ETOH_a_S19_uniq", "H2O_8h_ETOH_b_S20_uniq", "H2O_8h_ETOH_d_S21_uniq",
              "DOX_8h_ETOH_a_S22_uniq", "DOX_8h_ETOH_b_S23_uniq", "DOX_8h_ETOH_d_S24_uniq",
              "H2O_8h_4OHT_a_S25_uniq", "H2O_8h_4OHT_b_S26_uniq", "H2O_8h_4OHT_d_S27_uniq",
              "DOX_8h_4OHT_a_S28_uniq", "DOX_8h_4OHT_b_S29_uniq", "DOX_8h_4OHT_d_S30_uniq")]
```

```{r}
#row_clustering <- hclust(dist(mat[,c("H2O_8h_4OHT_a_S25_uniq", "H2O_8h_4OHT_b_S26_uniq", "H2O_8h_4OHT_d_S27_uniq",
#                                     "H2O_8h_ETOH_a_S19_uniq", "H2O_8h_ETOH_b_S20_uniq", "H2O_8h_ETOH_d_S21_uniq",
#                                     "DOX_8h_4OHT_a_S28_uniq", "DOX_8h_4OHT_b_S29_uniq", "DOX_8h_4OHT_d_S30_uniq")]))
```

Row annotation (gene annotation)
```{r}
row_anno <- data.frame(row.names = rownames(mat), pErk_DE = up_and_down_tam_vs_etoh$DE)

row_anno$pErk_mutant_DE <- "No change"
row_anno$pErk_mutant_DE[row.names(row_anno) %in% intersect(row.names(row_anno), up_dox_tam_vs_h2o_tam)] <- "Up"
row_anno$pErk_mutant_DE[row.names(row_anno) %in% intersect(row.names(row_anno), down_dox_tam_vs_h2o_tam) ] <- "Down"
```

Column annotation (sample annotation)
```{r}
anno <- as.data.frame(colData(dds)[c("H2O_8h_4OHT_a_S25_uniq", "H2O_8h_4OHT_b_S26_uniq", "H2O_8h_4OHT_d_S27_uniq",
                                     "DOX_8h_4OHT_a_S28_uniq", "DOX_8h_4OHT_b_S29_uniq", "DOX_8h_4OHT_d_S30_uniq",
                                     "H2O_8h_ETOH_a_S19_uniq", "H2O_8h_ETOH_b_S20_uniq", "H2O_8h_ETOH_d_S21_uniq", 
                                     "DOX_8h_ETOH_a_S22_uniq", "DOX_8h_ETOH_b_S23_uniq", "DOX_8h_ETOH_d_S24_uniq"), c("Dox","Tam")])
colnames(anno) <- c("Oct4", "pErk")
anno$Oct4 <- as.character(anno$Oct4)
anno$Oct4[which(anno$Oct4 == "DOX")] <- "-"
anno$Oct4[which(anno$Oct4 == "H2O")] <- "+"
anno$Oct4 <- as.factor(anno$Oct4)

anno$pErk <- as.character(anno$pErk)
anno$pErk[which(anno$pErk == "4OHT")] <- "+"
anno$pErk[which(anno$pErk == "ETOH")] <- "-"
anno$pErk <- as.factor(anno$pErk)
```

Annotation colors
```{r}
ann_colors <- list(Oct4 = c("+"="grey45", "-"= "grey75"), pErk = c("+"="grey45", "-"= "grey75"),
                   pErk_DE = c("Up" = "firebrick2", "Down" = "skyblue2"),
                   pErk_mutant_DE = c("Up" = "firebrick2", "Down" = "skyblue2", "No change" = "grey"))
```

Create heatmap
```{r original_heatmap, fig.height=7, fig.width=7}
hm<- pheatmap(#filename = "../results/tam_vs_etoh_annotated_heatmap.png"),  
              mat, annotation_col = anno, annotation_colors = ann_colors, 
              annotation_row = row_anno, 
              show_rownames = F, show_colnames = F, 
              cluster_cols = F, cluster_rows = T, 
              main = "Dysregulation of Fgf-Erk dependent genes\nAbsolute LFC >1, adj. p-value <0.05", scale = "row",
              treeheight_row = 0, treeheight_col = 0, width = 7, height = 9)
```

# Ordered plot
```{r}
ordered_mat <- mat[order(row_anno[,1],row_anno[,2],decreasing = T),]
```


```{r ordered_heatmap, fig.height=7, fig.width=7}
hm<- pheatmap(#filename = paste0(dir,"results/tam_vs_etoh_annotated_heatmap.png"),  
              ordered_mat, annotation_col = anno, annotation_colors = ann_colors, 
              annotation_row = row_anno, 
              show_rownames = F, show_colnames = F, 
              cluster_cols = F, cluster_rows = F, 
              main = "Dysregulation of Fgf-Erk dependent genes\nAbsolute LFC >1, adj. p-value <0.05", scale = "row",
              treeheight_row = 0, treeheight_col = 0, width = 7, height = 9)
```

# Save gene annotations (row_anno)
```{r}
row_anno$Geneid <- rownames(row_anno)

row_anno <- merge(row_anno, mcols(dds)[,c("Geneid","GeneSymbol")], by = "Geneid")
write.table(row_anno, file = paste0(results_folder,"tam_vs_etoh_annotated_heatmap_row_anno.tsv"), 
            row.names = F, col.names = T, sep = "\t", quote = F)
```


# Session info

```{r session-info, cache = FALSE}
devtools::session_info()
```
